Temporal Variations In Vapor Intrusion-Induced Indoor Air Contaminant Concentrations

### 2019-04-01 tex/main.tex
Minor edits

Long term vapor intrusion (VI) studies in both residential and larger commercial structures have raised concerns regarding significant observed transient behavior in indoor air contaminant concentrations\cite{u.s._environmental_protection_agency_oswer_2015,folkes_observed_2009,holton_temporal_2013,johnston_spatiotemporal_2014,hosangadi_high-frequency_2017,mchugh_recent_2017,u.s._environmental_protection_agency_assessment_2015}.
Such variations make it difficult for those charged with protecting human health to formulate a response should evaluation of the risk of exposure be based upon observed peak concentrations, or long-term averages, or something else?
There is even uncertainty within the VI community regarding how to best develop sampling strategies to address this problem\cite{u.s._environmental_protection_agency_oswer_2015,holton_temporal_2013,johnson_integrated_2016}.
What represents a reasonable sampling strategy for a particular site a single 8 hr sample?
Repeated 8 hr samples?
Month-long samples?
Continuous monitoring?

VI involves the migration of volatilizing contaminants from soil, groundwater or other subsurface sources into overlying structures.
The basic nature of VI has been understood for some time and it has been the subject of much study, but some aspects remain poorly understood, such as the causes of the sometimes observed large temporal transients in indoor air concentrations.
Results from a house operated by Arizona State University (ASU) near Hill AFB in Utah, an EPA experimental house in Indianapolis, IN and a large warehouse at the Naval Air Station (NAS) North Island, CA have all shown significant transient variations in indoor air contaminant concentrations.
All were outfitted with sampling and monitoring equipment that allowed tracking temporal variation in indoor air contaminant concentrations on time scales of hours.
All have shown that these concentrations vary significantly with time - orders of magnitude on the timescale of a day or days\cite{holton_evaluation_2015,guo_vapor_2015,hosangadi_high-frequency_2017}.

In one instance the source of the variation was clearly established during the study of the site.
At the ASU house a drain pipe (or “land drain”) connected to a sewer system was discovered beneath the house.
Careful isolation of this source led to a clear conclusion that this “preferential pathway” for contaminant vapor migration significantly contributed to observed indoor air contaminant levels and their fluctuations\cite{guo_vapor_2015,guo_identification_2015}.
While in this case the issue of a contribution from a preferential pathway was clearly resolved, what it left open was a question of whether existence of such a preferential pathway would always be expected to lead to large fluctuations in indoor air contaminant concentrations.

Similarly, a sewer pipe has recently been suggested to be a source of the contaminants found in the EPA Indianapolis house.
That site was also characterized by large indoor air contaminant concentration fluctuations\cite{mchugh_evidence_2017,u.s._environmental_protection_agency_assessment_2015}.
Sewer lines have been previously implicated as VI sources at several sites\cite{pennell_sewer_2013,mchugh_evidence_2017,roghani_occurrence_2018,riis_vapor_2010}.
A Danish study has estimated that roughly 20\% of all VI sites in central Denmark involve significant sewer VI pathways\cite{nielsen_remediation_2017}.
Thus while consideration of sewer or other preferential pathways is now part of normal good practice in VI site investigation\cite{u.s._environmental_protection_agency_oswer_2015}, it is still not known whether the existence of such pathways automatically means that large temporal fluctuations are necessarily to be expected.

In some of the above cited cases\cite{pennell_sewer_2013,riis_vapor_2010}, a sewer provided a pathway for direct entry of contaminant into the living space.
While potentially important in many instances, this scenario is not further considered here where the focus is on pathways that deliver contaminant via the soil beneath a structure.
It is, however, now known that even absent a preferential pathway, there may be significant transient variation in indoor air contaminant concentrations at VI sites\cite{folkes_observed_2009,brenner_results_2010,johnston_spatiotemporal_2014}.
One example is the site at NAS North Island at which no preferential pathways have been identified.
Instead, a building at this site is characterized by significant temporal variations in indoor-outdoor pressure differential\cite{hosangadi_high-frequency_2017}.
It is believed that this is the origin of the observed indoor air contaminant concentration fluctuations at that site.

This paper investigates the sources of the temporal variation in indoor air contaminant concentrations in both the presence and absence of preferential pathways.
In this work, the latter scenarios are referred to as “normal” VI scenarios, in which there is typically a groundwater source of the contaminant.
Specifically, we pose the question of just how much variation in indoor air contaminant concentration may be expected at such normal VI sites vs. those characterized by preferential pathways within the soil beneath the site.
The conditions required for preferential pathways to become significant contributors to temporal variations in indoor air contaminant concentrations are also explored, and the consequences for sampling strategies are discussed.

@@

To frame the question of just how much variability in indoor air contaminant concentrations is actually seen, field datasets have been analyzed.
For this purpose, datasets from the ASU house in Utah, the EPA Indianapolis site and North Island NAS were chosen for analysis.
Readers are referred to the original published works for details regarding data acquisition\cite{holton_evaluation_2015,guo_vapor_2015,holton_temporal_2013,hosangadi_high-frequency_2017,u.s._environmental_protection_agency_assessment_2015}.

The ASU house data were obtained over a period of several years.
During part of this time, controlled pressure method (CPM) tests were being conducted, in which the house was underpressurized to an extent greater than that characterizing “normal” house operation thus increasing VI potential\cite{mchugh_evaluation_2012,mchugh_recent_2017,holton_evaluation_2015}.
The period of CPM testing is thus excluded from the analysis.
Likewise, the existence of a preferential pathway at the ASU house needs to be considered in examining that dataset; during some of the testing at that site, this pathway was cut off, resulting in “normal” VI conditions in which the main source of contaminant was diffusion of contaminant vapor from an underlying groundwater source.

The NAS North Island dataset has not (as far as is known) been influenced by a preferential pathway, but the structure there was subject to “large” internal pressure fluctuations.
By “large” is meant still only of order 10-20 Pa, but these were greater than those generally recorded at the ASU house during normal operations.
The underlying soil at NAS North Island is sandy\cite{hosangadi_high-frequency_2017} and more permeable than that at the ASU site, which will be shown to lead to greater pressure sensitivity in the former case.

The Indianapolis site investigation also spanned a number of years and periodically included the testing of a sub-slab depressurziation system (SSD) for VI mitigation.
Only the period before the installation of this system was considered in the present analysis.
It is likely a sewer line beneath the structure acted as a preferential pathway\cite{mchugh_evidence_2017}.
Unlike at the ASU house, this preferential pathway was never removed or blocked, making it impossible to isolate the role of the preferential pathway at this site.
It is still of interest to consider the data from this site because of the completeness and extensiveness of the data collection.
Figure \ref{fig:indianapolis} illustrates a typical reported series of indoor air trichloroethylene (TCE) concentration measurements from this site.
There is almost a two order of magnitude variation in the concentration data.

Some of the analysis of the above three field data sets relies on a probability density estimation technique called ”kernel density estimation” (KDE).
KDE is a technique used for estimating the probability distribution of a random variable(s) by using multiple kernels, or weighting functions to characterize the data sets.
In this case, Gaussian kernels are used to create the KDEs.
This means that it is presumed that the variables of interest (i.e., indoor air contaminant concentrations and indoor-outdoor pressure differentials, as sampled) are normally distributed around mean values and that there are statistical fluctuations associated with each sampling event.
In this instance, the scipy statistical package was used to construct the KDEs, assuming a bandwidth parameter determined by Scott’s rue.
The SciPy Python library was used to conduct all statistical analysis and data processing\cite{jones_scipy_2011}.

@@

High frequency measurement of indoor air contaminant concentrations, $c_\mathrm{in}$ , such as those in Figure 1, took place at both the ASU House and the Indianapolis House over significant periods\cite{u.s._environmental_protection_agency_assessment_2015,holton_temporal_2013}.
Furthermore, at the Indianapolis site the $c_\mathrm{in}$ for three different contaminants, chloroform, TCE, and tetrachloroethylene (PCE) were all collected, allowing examination of the variability of each VI contaminant.
The NAS North Island NAS dataset was obtained over a much shorter duration, and is therefore not examined in this portion of the analysis.

Figure \ref{fig:indianapolis} showed a large degree of temporal variation in one of the components, and the data for the other components were quite similar.
What is apparent upon closer examination of such data is that the actual day-to-day variations are typically not nearly as large as those observed when tracking the data for a longer time.
To demonstrate this point, the quotient of the maximum and minimum $c_\mathrm{in}$ values (denoted as $c_\mathrm{max}/c_\mathrm{min}$) are shown as a function of time in Figure \ref{fig:resampling}.
The values shown in Figure \ref{fig:resampling} are the means of the quotients calculated for samples separated by the indicated times and the error bars indicate the 95th percentile of all the data points.
Hypothetical resampling periods of one, two, three days, and the same number of weeks, and months were chosen.

For example, if the data are examined in terms of the mean maximum variation observable over the course of 24 hours (one day) the variation is no greater than about a factor of two for any of the contaminants at the Indianapolis house or for TCE at the ASU house (when the preferential pathway was closed).
The mean variability at the latter was only a bit higher (about a factor of 3) when the preferential pathway was open.
In other words, a sampling protocol that involves sampling on two consecutive days would typically not uncover the large temporal variations that characterize the site over longer periods of time.
As Figure \ref{fig:indianapolis} shows, there are certainly isolated days in which a larger daily change was observed, but these were not typical, to the extent that they fall outside of the 95\% criteria used in defining the error bars.
So while such unusual jumps might be seen (for unknown reasons) in a very small percentage of cases, the expectation is much more represented by what is shown in Figure \ref{fig:resampling}.

Weeks of temporal separation in sampling events are required to observe the large variations of concern.
Orders of magnitude differences begin to manifest themselves over the course of months.
This is not surprising, since those who performed the measurements have already reported that there were seasonal aspects to the values obtained.
This would be consistent with requiring months to see the more significant variations.

This analysis also suggests that certain types of preferential pathways contribute to larger variations on shorter timescales (ASU House).
Even though there was a preferential pathway present at the Indianapolis House, the transients associated with its presence were of a slower nature and the behavior was not unlike what was observed at ASU House when the preferential pathway was closed.
This warns that the mere existence of a preferential pathway is not by itself sufficient to create a situation of large variations over short sampling times.

The longer the resampling period, the larger the maximum variability in observed indoor air contaminant concentrations.
In the case of the ASU House with the preferential pathway open, the variability went from less than a threefold difference on the timescale of a day, to two to three orders of magnitude over the course of weeks.
Thus there are different timescales that characterize different extents of variation, again pointing to the existence of more than a single factor that determines variability.

If the maximum observed variability is quite small across only a few days as these data suggest, a sampling strategy that involves taking daily samples for several days in a row will likely yield little extra useful information beyond what can be obtained with a single sample.

@@

The pressure difference between the indoor and outdoor/ambient ($p_\mathrm{in/out}$) leads to advection, by which contaminants are drawn into (or prevented from) entering a structure.  Changes in $p_\mathrm{in/out}$ can take place quickly, leaving open the possibility of their impacting VI far more rapidly than can fluctuations in say groundwater depth or contaminant concentration (these latter processes take weeks or even months to impact the overlying structure).
We examine the relationship between $p_\mathrm{in/out}$ and $c_\mathrm{in}$ by constructing the two-dimensional kernel density estimation (KDE) plots seen in Figure \ref{fig:kde}.
The KDE plots allow us to view the measured distributions of $p_\mathrm{in/out}$ and $c_\mathrm{in}$, and develop a visual impression of how well these distributions correlate with one another.
For this analysis we considered two VI sites, NAS North Island and the ASU House.
The ASU House dataset was divided into two periods, one before and the other after the land drain (called the preferential pathway (PP) from  ere on) had been closed.
By comparing these two periods on a single plot, the impact of the preferential pathway becomes clearer.\par

In Figure \ref{fig:kde}, the indoor air contaminant concentration $c_\mathrm{in}$ is normalized to the mean $c_\mathrm{in,mean}$ of each dataset, allowing comparison of the impact $p_\mathrm{in/out}$ on $c_\mathrm{in}$ independently from the large differences in absolute values of indoor air concentrations at the different sites.
A value of 10 on the y-axis indicates that the corresponding plotted value of $c_\mathrm{in}$ is 10 times greater than the mean for the dataset, and 0.1 indicate that it is one tenth of the mean.
Inspection of the range of normalized $c_\mathrm{in}$ values in Figure \ref{fig:kde} again shows the two order of magnitude spread in observed values, implying a sampling at one particular time might give a value that is two orders of magnitude different than a result from a different time.
Such issues have of course already been pointed out by the investigators who obtained the data.

The power of this KDE representation is that it permits evaluation of the relationship of two independently measured data - the indoor air contaminant concentration and the indoor-outdoor pressure difference.
Examining the data in this manner immediately points to an important difference between the data from the ASU House and those from NAS North Island.
At NAS North Island site $p_\mathrm{in/out}$ varies significantly; the 5th and 95th percentile of $p_\mathrm{in/out}$ are -19.9 and 7.4 Pa respectively.
This may be contrasted with 5th and 95th percentile $p_\mathrm{in/out}$ at the ASU house: -1.4 and 2.1 Pa (with the PP open), and -2.1 and 2.27 Pa (PP closed).

The much larger under- and overpressurization of the NAS North Island site compared to the ASU House makes the pressure dependence of indoor air concentration much more visible at the former site.
The Pearson’s r value for the correlation between $p_\mathrm{in/out}$ and $c_\mathrm{in}$ for each dataset is shown in the legend, and confirms what is apparent to the eye; the pressure driving force is a determining factor for observed contamination at NAS North Island.
But the broadness of the band of the NAS North Island concentration data set suggests that there is still a source of variability in $c_\mathrm{in}$ that has not been fully captured - this will be addressed below.

The ASU house datasets offer a different picture. The variability   of $c_\mathrm{in}$ is just as large, or even larger than at NAS North Island, yet the $p_\mathrm{in/out}$ varied far less. The weaker dependence of $c_\mathrm{in}$ on the pressure difference is confirmed by the much lower r values for the correlations between the variables. In other words, there is not nearly as strong a correlation between variation in indoor air contaminant concentration and pressure difference for the ASU House as there was for NAS North Island.

It should be noted that control of indoor depressurization is the principle that underlies the controlled pressure method (CPM) for assessing the potential severity of a VI situation.
These results strongly suggest that there are other factors besides indoor pressure determining indoor air contaminant concentrations, and their variations, that may not be accounted for in applying this method.

The data for the ASU House also offer an insight into the role of the preferential pathway.
At first glance it may seem like the $c_\mathrm{in}$ values for the periods when the PP is open and closed are relatively comparable.
However, the 5th and 95th percentiles values of $c_\mathrm{in}$/$c_\mathrm{in,mean}$ differ significantly as may be seen in Table \ref{tbl:percentiles}.
It is clear that existence of the preferential pathway dramatically increases the variability in indoor air contaminant concentration.
This again is entirely consistent with what the investigators of that site have already reported\cite{guo_identification_2015}.
The correlation with indoor-outdoor pressure difference is weak in the ASU house cases, so there are clearly factors other than pressure difference that determine the variability in each.
These will be explored with the help of a modeling analysis presented below.

@@

Observed temporal variations in indoor air contaminant concentrations might be explained by temporal variations in subslab contaminant concentrations.
To examine how variability in subslab contaminant concentration might contribute to variability in indoor air contaminant concentration, data on the attenuation from subslab ($\alpha_\mathrm{subslab} = c_\mathrm{in}/c_\mathrm{subslab}$) were examined.
The dataset utilized for this was that from the ASU House.
The $c_\mathrm{subslab}$ values were taken from a soil gas probe labeled as "6" at the ASU house.
This probe was located closest to both the exit of the preferential pathway pipe, and to a reported breach in the foundation that served as a key entry pathway for contaminant getting into the house\cite{guo_identification_2015}.
The results are shown in Figure \ref{fig:attenuation_subslab}.

It is apparent that during the period when the preferential pathway was closed, $\alpha_\mathrm{subslab}$ did not vary significantly, and was quite close to the EPA recommended $\alpha_\mathrm{subslab}$ value of 0.03\cite{u.s._environmental_protection_agency_oswer_2015}.
Thus during the period when the preferential pathway was closed large temporal variations in subslab concentrations could not have been driving the variations in indoor air contaminant concentrations.

When the PP was open, there was considerably more variability in the subslab concentration values, and the mean value was higher than in the case where the preferential pathway was closed.
It was also not uncommon for the observed $\alpha_\mathrm{subslab}$ to exceed unity.
While large $\alpha_\mathrm{subslab}$ values may sometimes indicate indoor sources at a site, there were none at the ASU house.
A more likely explanation is that even though probe "6" was located in close proximity to the exit of the preferential pathway, there might have still existed significant spatial variability in $c_\mathrm{subslab}$ that could not be captured with a single measurement.
This suggests caution is needed in profiling subslab contaminant concentrations in the presence of preferential pathways - significant variations are possible.

What the results of Figure \ref{fig:attenuation_subslab} do clearly show is that the existence of a preferential pathway of the kind at ASU House (and idealized in Figure \ref{fig:indianapolis}) can influence the temporal variation of subslab concentrations in a much less predictable way than those observed in “normal” VI scenarios.

@@

To begin a more thorough discussion regarding the role and impact a preferential pathway may have at a particular site, we turn to our model presented in the methods section.
Here, $A_e$ is assumed to be a constant 0.5 per hour, and $p_\mathrm{in/out}$ is varied from -5 to 5 Pa.
The predicted $c_\mathrm{in}$ is then normalized to the groundwater concentration $c_\mathrm{gw}$, giving the attenuation from groundwater $\alpha_\mathrm{gw}$.

The predicted $\alpha_\mathrm{gw}$ as a function of $p_\mathrm{in/out}$ is given by the central blue line in the upper panel of Figure \ref{fig:land_drain_scenarios} (ignore the shaded blue areas for now).
These predicted values are compared to $\alpha_\mathrm{gw}$ from the ASU house data from the period the preferential pathway was open, given by the blue dots.

>>>>

Long term vapor intrusion (VI) studies in both residential and larger commercial structures have raised concerns regarding significant observed transient behavior in indoor air contaminant concentrations\cite{u.s._environmental_protection_agency_oswer_2015,folkes_observed_2009,holton_temporal_2013,johnston_spatiotemporal_2014,hosangadi_high-frequency_2017,mchugh_recent_2017,u.s._environmental_protection_agency_assessment_2015}.
Such variations make it difficult for those charged with protecting human health to formulate a response should evaluation of the risk of exposure be based upon observed peak concentrations, or long-term averages, or something else?
There is even uncertainty within the VI community regarding how to best develop sampling strategies to address this problem\cite{u.s._environmental_protection_agency_oswer_2015,holton_temporal_2013,johnson_integrated_2016}.
What represents a reasonable sampling strategy for a particular site a single 8 hr sample?
Repeated 8 hr samples?
Month-long samples?
Continuous monitoring?\par

VI involves the migration of volatilizing contaminants from soil, groundwater or other subsurface sources into overlying structures.
The basic nature of VI has been understood for some time and it has been the subject of much study, but some aspects remain poorly understood, such as the causes of the sometimes observed large temporal transients in indoor air concentrations.
Results from a house operated by Arizona State University (ASU) near Hill AFB in Utah, an EPA experimental house in Indianapolis, IN and a large warehouse at the Naval Air Station (NAS) North Island, CA have all shown significant transient variations in indoor air contaminant concentrations.
All were outfitted with sampling and monitoring equipment that allowed tracking temporal variation in indoor air contaminant concentrations on time scales of hours.
All have shown that these concentrations vary significantly with time - orders of magnitude on the timescale of a day or days\cite{holton_evaluation_2015,guo_vapor_2015,hosangadi_high-frequency_2017}.\par

In one instance the source of the variation was clearly established during the study of the site.
At the ASU house a drain pipe (or “land drain”) connected to a sewer system was discovered beneath the house.
Careful isolation of this source led to a clear conclusion that this “preferential pathway” for contaminant vapor migration significantly contributed to observed indoor air contaminant levels and their fluctuations\cite{guo_vapor_2015,guo_identification_2015}.
While in this case the issue of a contribution from a preferential pathway was clearly resolved, what it left open was a question of whether existence of such a preferential pathway would always be expected to lead to large fluctuations in indoor air contaminant concentrations.\par

Similarly, a sewer pipe has recently been suggested to be a source of the contaminants found in the EPA Indianapolis house.
That site was also characterized by large indoor air contaminant concentration fluctuations\cite{mchugh_evidence_2017,u.s._environmental_protection_agency_assessment_2015}.
Sewer lines have been previously implicated as VI sources at several sites\cite{pennell_sewer_2013,mchugh_evidence_2017,roghani_occurrence_2018,riis_vapor_2010}.
A Danish study has estimated that roughly 20\% of all VI sites in central Denmark involve significant sewer VI pathways\cite{nielsen_remediation_2017}.
Thus while consideration of sewer or other preferential pathways is now part of normal good practice in VI site investigation\cite{u.s._environmental_protection_agency_oswer_2015}, it is still not known whether the existence of such pathways automatically means that large temporal fluctuations are necessarily to be expected.\par

In some of the above cited cases\cite{pennell_sewer_2013,riis_vapor_2010}, a sewer provided a pathway for direct entry of contaminant into the living space.
While potentially important in many instances, this scenario is not further considered here where the focus is on pathways that deliver contaminant via the soil beneath a structure.
It is, however, now known that even absent a preferential pathway, there may be significant transient variation in indoor air contaminant concentrations at VI sites\cite{folkes_observed_2009,brenner_results_2010,johnston_spatiotemporal_2014}.
One example is the site at NAS North Island at which no preferential pathways have been identified.
Instead, a building at this site is characterized by significant temporal variations in indoor-outdoor pressure differential\cite{hosangadi_high-frequency_2017}.
It is believed that this is the origin of the observed indoor air contaminant concentration fluctuations at that site.\par

This paper investigates the sources of the temporal variation in indoor air contaminant concentrations in both the presence and absence of preferential pathways.
In this work, the latter scenarios are referred to as “normal” VI scenarios, in which there is typically a groundwater source of the contaminant.
Specifically, we pose the question of just how much variation in indoor air contaminant concentration may be expected at such normal VI sites vs. those characterized by preferential pathways within the soil beneath the site.
The conditions required for preferential pathways to become significant contributors to temporal variations in indoor air contaminant concentrations are also explored, and the consequences for sampling strategies are discussed.\par

@@

To frame the question of just how much variability in indoor air contaminant concentrations is actually seen, field datasets have been analyzed.
For this purpose, datasets from the ASU house in Utah, the EPA Indianapolis site and North Island NAS were chosen for analysis.
Readers are referred to the original published works for details regarding data acquisition\cite{holton_evaluation_2015,guo_vapor_2015,holton_temporal_2013,hosangadi_high-frequency_2017,u.s._environmental_protection_agency_assessment_2015}.\par

The ASU house data were obtained over a period of several years.
During part of this time, controlled pressure method (CPM) tests were being conducted, in which the house was underpressurized to an extent greater than that characterizing “normal” house operation thus increasing VI potential\cite{mchugh_evaluation_2012,mchugh_recent_2017,holton_evaluation_2015}.
The period of CPM testing is thus excluded from the analysis.
Likewise, the existence of a preferential pathway at the ASU house needs to be considered in examining that dataset; during some of the testing at that site, this pathway was cut off, resulting in “normal” VI conditions in which the main source of contaminant was diffusion of contaminant vapor from an underlying groundwater source.\par

The NAS North Island dataset has not (as far as is known) been influenced by a preferential pathway, but the structure there was subject to “large” internal pressure fluctuations.
By “large” is meant still only of order 10-20 Pa, but these were greater than those generally recorded at the ASU house during normal operations.
The underlying soil at NAS North Island is sandy\cite{hosangadi_high-frequency_2017} and more permeable than that at the ASU site, which will be shown to lead to greater pressure sensitivity in the former case.\par

The Indianapolis site investigation also spanned a number of years and periodically included the testing of a sub-slab depressurziation system (SSD) for VI mitigation.
Only the period before the installation of this system was considered in the present analysis.
It is likely a sewer line beneath the structure acted as a preferential pathway\cite{mchugh_evidence_2017}.
Unlike at the ASU house, this preferential pathway was never removed or blocked, making it impossible to isolate the role of the preferential pathway at this site.
It is still of interest to consider the data from this site because of the completeness and extensiveness of the data collection.
Figure \ref{fig:indianapolis} illustrates a typical reported series of indoor air trichloroethylene (TCE) concentration measurements from this site.
There is almost a two order of magnitude variation in the concentration data.\par

Some of the analysis of the above three field data sets relies on a probability density estimation technique called ”kernel density estimation” (KDE).
KDE is a technique used for estimating the probability distribution of a random variable(s) by using multiple kernels, or weighting functions to characterize the data sets.
In this case, Gaussian kernels are used to create the KDEs.
This means that it is presumed that the variables of interest (i.e., indoor air contaminant concentrations and indoor-outdoor pressure differentials, as sampled) are normally distributed around mean values and that there are statistical fluctuations associated with each sampling event.
In this instance, the scipy statistical package was used to construct the KDEs, assuming a bandwidth parameter determined by Scott’s rue.
The SciPy Python library was used to conduct all statistical analysis and data processing\cite{jones_scipy_2011}.\par

@@

High frequency measurement of indoor air contaminant concentrations, $c_\mathrm{in}$ , such as those in Figure 1, took place at both the ASU House and the Indianapolis House over significant periods\cite{u.s._environmental_protection_agency_assessment_2015,holton_temporal_2013}.
Furthermore, at the Indianapolis site the $c_\mathrm{in}$ for three different contaminants, chloroform, TCE, and tetrachloroethylene (PCE) were all collected, allowing examination of the variability of each VI contaminant.
The NAS North Island NAS dataset was obtained over a much shorter duration, and is therefore not examined in this portion of the analysis.\par

Figure \ref{fig:indianapolis} showed a large degree of temporal variation in one of the components, and the data for the other components were quite similar.
What is apparent upon closer examination of such data is that the actual day-to-day variations are typically not nearly as large as those observed when tracking the data for a longer time.
To demonstrate this point, the quotient of the maximum and minimum $c_\mathrm{in}$ values (denoted as $c_\mathrm{max}/c_\mathrm{min}$) are shown as a function of time in Figure \ref{fig:resampling}.
The values shown in Figure \ref{fig:resampling} are the means of the quotients calculated for samples separated by the indicated times and the error bars indicate the 95th percentile of all the data points.
Hypothetical resampling periods of one, two, three days, and the same number of weeks, and months were chosen.\par

For example, if the data are examined in terms of the mean maximum variation observable over the course of 24 hours (one day) the variation is no greater than about a factor of two for any of the contaminants at the Indianapolis house or for TCE at the ASU house (when the preferential pathway was closed).
The mean variability at the latter was only a bit higher (about a factor of 3) when the preferential pathway was open.
In other words, a sampling protocol that involves sampling on two consecutive days would typically not uncover the large temporal variations that characterize the site over longer periods of time.
As Figure \ref{fig:indianapolis} shows, there are certainly isolated days in which a larger daily change was observed, but these were not typical, to the extent that they fall outside of the 95\% criteria used in defining the error bars.
So while such unusual jumps might be seen (for unknown reasons) in a very small percentage of cases, the expectation is much more represented by what is shown in Figure \ref{fig:resampling}.\par

Weeks of temporal separation in sampling events are required to observe the large variations of concern.
Orders of magnitude differences begin to manifest themselves over the course of months.
This is not surprising, since those who performed the measurements have already reported that there were seasonal aspects to the values obtained.
This would be consistent with requiring months to see the more significant variations.\par

This analysis also suggests that certain types of preferential pathways contribute to larger variations on shorter timescales (ASU House).
Even though there was a preferential pathway present at the Indianapolis House, the transients associated with its presence were of a slower nature and the behavior was not unlike what was observed at ASU House when the preferential pathway was closed.
This warns that the mere existence of a preferential pathway is not by itself sufficient to create a situation of large variations over short sampling times.\par

The longer the resampling period, the larger the maximum variability in observed indoor air contaminant concentrations.
In the case of the ASU House with the preferential pathway open, the variability went from less than a threefold difference on the timescale of a day, to two to three orders of magnitude over the course of weeks.
Thus there are different timescales that characterize different extents of variation, again pointing to the existence of more than a single factor that determines variability.\par

If the maximum observed variability is quite small across only a few days as these data suggest, a sampling strategy that involves taking daily samples for several days in a row will likely yield little extra useful information beyond what can be obtained with a single sample.\par

@@

The pressure difference between the indoor and outdoor/ambient ($p_\mathrm{in/out}$) leads to advection, by which contaminants are drawn into (or prevented from) entering a structure. 
Changes in $p_\mathrm{in/out}$ can take place quickly, leaving open the possibility of their impacting VI far more rapidly than can fluctuations in say groundwater depth or contaminant concentration (these latter processes take weeks or even months to impact the overlying structure).\par

We examine the relationship between $p_\mathrm{in/out}$ and $c_\mathrm{in}$ by constructing the two-dimensional kernel density estimation (KDE) plots seen in Figure \ref{fig:kde}.
The KDE plots allow us to view the measured distributions of $p_\mathrm{in/out}$ and $c_\mathrm{in}$, and develop a visual impression of how well these distributions correlate with one another.
For this analysis we considered two VI sites, NAS North Island and the ASU House.
The ASU House dataset was divided into two periods, one before and the other after the land drain (called the preferential pathway (PP) from  ere on) had been closed.
By comparing these two periods on a single plot, the impact of the preferential pathway becomes clearer.\par

In Figure \ref{fig:kde}, the indoor air contaminant concentration $c_\mathrm{in}$ is normalized to the mean $c_\mathrm{in,mean}$ of each dataset, allowing comparison of the impact $p_\mathrm{in/out}$ on $c_\mathrm{in}$ independently from the large differences in absolute values of indoor air concentrations at the different sites.
A value of 10 on the y-axis indicates that the corresponding plotted value of $c_\mathrm{in}$ is 10 times greater than the mean for the dataset, and 0.1 indicate that it is one tenth of the mean.\par

Inspection of the range of normalized $c_\mathrm{in}$ values in Figure \ref{fig:kde} again shows the two order of magnitude spread in observed values, implying a sampling at one particular time might give a value that is two orders of magnitude different than a result from a different time.
Such issues have of course already been pointed out by the investigators who obtained the data.\par

The power of this KDE representation is that it permits evaluation of the relationship of two independently measured data - the indoor air contaminant concentration and the indoor-outdoor pressure difference.
Examining the data in this manner immediately points to an important difference between the data from the ASU House and those from NAS North Island.
At NAS North Island site $p_\mathrm{in/out}$ varies significantly; the 5th and 95th percentile of $p_\mathrm{in/out}$ are -19.9 and 7.4 Pa respectively.
This may be contrasted with 5th and 95th percentile $p_\mathrm{in/out}$ at the ASU house: -1.4 and 2.1 Pa (with the PP open), and -2.1 and 2.27 Pa (PP closed).\par

The much larger under- and overpressurization of the NAS North Island site compared to the ASU House makes the pressure dependence of indoor air concentration much more visible at the former site.
The Pearson’s r-value for the correlation between $p_\mathrm{in/out}$ and $c_\mathrm{in}$ for each dataset is shown in the legend, and confirms what is apparent to the eye; the pressure driving force is a determining factor for observed contamination at NAS North Island.
But the broadness of the band of the NAS North Island concentration data set suggests that there is still a source of variability in $c_\mathrm{in}$ that has not been fully captured - this will be addressed below.\par

The ASU house datasets offer a different picture.
The variability of $c_\mathrm{in}$ is just as large, or even larger than at NAS North Island, yet the $p_\mathrm{in/out}$ varied far less.
The weaker dependence of $c_\mathrm{in}$ on the pressure difference is confirmed by the much lower r-values for the correlations between the variables.
In other words, there is not nearly as strong a correlation between variation in indoor air contaminant concentration and pressure difference for the ASU House as there was for NAS North Island.\par

It should be noted that control of indoor depressurization is the principle that underlies the controlled pressure method (CPM) for assessing the potential severity of a VI situation.
These results strongly suggest that there are other factors besides indoor pressure determining indoor air contaminant concentrations, and their variations, that may not be accounted for in applying this method.\par

The data for the ASU House also offer an insight into the role of the preferential pathway.
At first glance it may seem like the $c_\mathrm{in}$ values for the periods when the PP is open and closed are relatively comparable.
However, the 5th and 95th percentiles values of $c_\mathrm{in}$/$c_\mathrm{in,mean}$ differ significantly as may be seen in Table \ref{tbl:percentiles}.
It is clear that existence of the preferential pathway dramatically increases the variability in indoor air contaminant concentration.
This again is entirely consistent with what the investigators of that site have already reported\cite{guo_identification_2015}.
The correlation with indoor-outdoor pressure difference is weak in the ASU house cases, so there are clearly factors other than pressure difference that determine the variability in each.
These will be explored with the help of a modeling analysis presented below.\par

@@

Observed temporal variations in indoor air contaminant concentrations might be explained by temporal variations in subslab contaminant concentrations.
To examine how variability in subslab contaminant concentration might contribute to variability in indoor air contaminant concentration, data on the attenuation from subslab ($\alpha_\mathrm{subslab} = c_\mathrm{in}/c_\mathrm{subslab}$) were examined.
The dataset utilized for this was that from the ASU House.
The $c_\mathrm{subslab}$ values were taken from a soil gas probe labeled as "6" at the ASU house.
This probe was located closest to both the exit of the preferential pathway pipe, and to a reported breach in the foundation that served as a key entry pathway for contaminant getting into the house\cite{guo_identification_2015}.
The results are shown in Figure \ref{fig:attenuation_subslab}.\par

It is apparent that during the period when the preferential pathway was closed, $\alpha_\mathrm{subslab}$ did not vary significantly, and was quite close to the EPA recommended $\alpha_\mathrm{subslab}$ value of 0.03\cite{u.s._environmental_protection_agency_oswer_2015}.
Thus during the period when the preferential pathway was closed large temporal variations in subslab concentrations could not have been driving the variations in indoor air contaminant concentrations.\par

When the PP was open, there was considerably more variability in the subslab concentration values, and the mean value was higher than in the case where the preferential pathway was closed.
It was also not uncommon for the observed $\alpha_\mathrm{subslab}$ to exceed unity.
While large $\alpha_\mathrm{subslab}$ values may sometimes indicate indoor sources at a site, there were none at the ASU house.
A more likely explanation is that even though probe "6" was located in close proximity to the exit of the preferential pathway, there might have still existed significant spatial variability in $c_\mathrm{subslab}$ that could not be captured with a single measurement.
This suggests caution is needed in profiling subslab contaminant concentrations in the presence of preferential pathways - significant variations are possible.\par

What the results of Figure \ref{fig:attenuation_subslab} do clearly show is that the existence of a preferential pathway of the kind at ASU House (and idealized in Figure \ref{fig:indianapolis}) can influence the temporal variation of subslab concentrations in a much less predictable way than those observed in “normal” VI scenarios.\par

@@

Having established the potential impacts of certain inputs on determining variability in indoor air contaminant concentrations, the mathematical model of VI can help further elucidate other key aspects.
The results of calculations on a scenario corresponding to Figure \ref{fig:model} are presented in Figure \ref{fig:land_drain_scenarios}.
This scenario is not intended to exactly represent the situation at ASU House, but it is similar in the key aspect of having a preferential pathway delivering contaminant to a gravel sub-base.
The full, complex geometry of the ASU House has not been represented, but the modeled structure is of comparable size, and will be subject to operational parameters based upon what were measured at that site.
The general modeling conditions are those shown in Table \ref{fig:indianapolis}.\par

The model successfully predicts the observed trends in $\alpha_\mathrm{gw}$ as $p_\mathrm{in/out}$ decreases (increased depressurization) but somewhat underpredicts $\alpha_\mathrm{gw}$ as the house is overpressurized.
Most significantly, the model captures that even for a small increase in depressurization (0 to -5 Pa) a very large increase in $\alpha_\mathrm{gw}$ (two order of magnitude) can occur.\par

The asymmetry relative to the predictions for depressurization and overpressurization is due to two factors.
First, the preferential pathway acts not only as a source of contaminant vapor, but also  as a source of air to the subslab.
Because of the large resistance to soil gas flow in the surrounding soil, having a local source of air to support the increase of advective flow into the structure from the subslab region makes a large difference.\par
